What Can Stack Traces Tell us about the Use of Exceptions in Java?

### 2014-05-07 android-stacks.tex
reducing backgounf section and introduction

To guide this exploration we compiled general guidelines on how to use
exceptions proposed by Gosling~\cite{gosling2000java},
Wirfs-Brock~\cite{wirfs2006toward} and Bloch~\cite{bloch2008effective} and then
we investigate what stack traces can tell us about the adherence to such
practices in Java programs.
%~\todo{This should be our motivation, as this is
%our biggest contribution. I already changed the text. Lets discuss about this
%new version}

%GGG: I think this is too much detail for the introduction In the second phase
%of this study, we combined the information of the mined stack traces with
%source code and bytecode analysis. To do so we selected the subset of 482
%Android repositories. The Android repositories were chosen because this kind of
%application is known for dealing with several sources of exceptions, as they
%are usually io-intensive (i.e., deals with Network, Bluetooth, camera, and
%other resources), and multi-htreaded. Moreover, they are based on a single
%platform which allowed more fine-grained analysis of their exception behaviour.

@@

%GGG: Personally, I hate this paragraph in all papers as it usually reads like a space-filler
The remainder of this paper is organized as follows. Section 2 presents a
background on error handling mechanisms and Java exception handling constructs
and practices. Section 3 defines research questions, the configuration of our
exploratory study, the target applications and the evaluation procedures.
Section 4 reports our analysis of the empirical data collected relating to the
first research questions. Section X is dedicated to provide insights for the 3rd
research question. Section 6 provides further discussions and lessons learned
and discusses the threats to validity associated to this study. Finally Section
7 describes the related work, and Section 8 presents our conclusions and
directions for future work.

@@

\subsection{Error handling in programming languages}

Techniques for error detection and handling are no more an optional add-on but a
fundamental part of the system~\cite{bruntink2006discovering}. Several
approaches exists for detecting errors during system execution and responding to
them accordingly, such as the return-code idiom (very popular in C programs and
operating systems), and the exception handling mechanism, embedded in many main
stream programming languages.

@@

In Java, an exception is represented according to the class hierarchy shown in
Figure~\ref{fig:exchier}.  According to this structure every exception is an
instance of the Throwable class, and can be of three kinds: checked exceptions
(extends Exception), run-time exceptions (extends RuntimeException), and errors
(extends Error)~\cite{gosling2000java}.

@@

In Java an exception can be signaled in one of the following
circumstances~\cite{gosling2000java}:

\begin{itemize} \item  explicitly thrown when a throw statement is executed;
  \item  implicitly thrown by the JVM when the evaluation of an expression
    violates the normal semantics of language (e.g., out-of-bounds array index,
    division-by-zero, access to a null reference); or \item implicitly thrown by
    the JVM due to an internal error or resource limitation (e.g.,
    OutofMemoryError).  \end{itemize}
%\todo{This sub-section is quite similar to the section introduction, shouldn't
%we merge them?}
When an exception is signaled it causes a transfer of control from the point
where the exception occurred to a point that can be specified by the programmer
(exception handler). In Java it is represented by the try-catch block and is
responsible for performing any recovery action necessary to bring the software
system back to a normal state and, whenever this is not possible, to log the
exception and abort the system in an expected safe way.

In Java, as in many languages, the search for the handler to deal with a raised
exception occurs along the dynamic invocation chain. This is claimed to increase
the reusability, since the invoker of an operation can handle it in a wider
context ~\cite{goodenough1975exception}. If no handler is found in the chain,
the default handler associated to the Thread or ThreadGroup is invoked, which
usually terminates the system and prints the exception stack trace.

\subsubsection{Exception Chaining} Exception Chaining is a common way of
propagating exceptions across components of different abstraction levels. It is
implemented by special exception constructor which receives another exception as
a parameter. An example a multiple exception chaining is depicted on the stack
trace in Figure~\ref{fig:wrapping}.
%This stack trace was extracted from an issue of an Android application in
%Github.

@@

Such undocumented runtime exceptions represent a serious threat to system robustness, specially
when such exceptions are thrown by reused third party code (e.g. libraries, or framework utility code).
Since it is very difficult or even impossible for the client of such thrid party code to 
protect the application against undocumented runtime exceptions, the
 undocumented runtime exception may remain uncaught and lead to  system crashes.

@@

 Manually inspecting the library code of methods signaling runtime exceptions 
 - this analysis did not consider the exceptions implicitly signaled  by JVM (e.g. NullPoiterException) -
 we could also observe that none of the runtime exceptions signaled by them were documented.
Examples of such methos are listed on the Table X.

This result is aligned with the results of other study conducted by from Cabral
and Marcques ~\cite{} which ialso observe that runtime exceptions in .Net programs 
are most often not documented.

On the other hand, we could observe that most of the exceptions signaled by the
set of libraries reused by Android platform (i.e., org.apache.harmony,
org.w3c.dom, sun.misc, org.apache.http, org.json, org.xml, javax.) referred in
this work as libcore, were checked exceptions. It is pointed as a good practice
for libraries (see Section ~\ref{best}) because by using checked exceptions
libraries can define a precise exception interface to its clients. Such
libraries are heavily used in several projects, and such precise exception
interface may be related to the libraries maturity.

%\textbf{

% the developer does not have access to their
%source code and thus needs to rely on the library documentation about the
%runtime exceptions that should be thrown – which, more often than not, are
%neither complete nor precise (Thomas, 2002; Sacramento et al., 2006). As a
%consequence, unhandled unchecked exceptions can be seen as one of the major
%sources of bugs in current Java systems (Jo et al., 2004). }

%On the other hand, we could observe that most of the exceptions signaled by the
%set of libraries reused by Android platform (i.e., org.apache.harmony,
%org.w3c.dom, sun.misc, org.apache.http, org.json, org.xml, javax.) referred in
%this work as libcore, were checked exceptions. It is pointed as a good practice
%for libraries (see Section ~\ref{best}) because by using checked exceptions
%libraries can define a precise exception interface to its clients. Such
%libraries are heavily used in several projects, and such precise exception
%interface may be related to the libraries maturity.

%Manually inspecting the code signalers which RuntimeExceptions were thrown in Android
%Platform we could observe that direct instances of java.lang.RuntimeException
% were thrown due to two main reasons: (i) when there was no recovery action associated
%to the exceptional condition, which means that regardless the kind of exception the
%only possible action was to abort the application; and when translating exceptions
%signaled by application methods. Since Android platform is heavily based on
%call backs", when a application throws a runtime exception (i.e., one of the life cicle 
%methods of an Activity) and does it, the exception scapes to compotenst of the Android platform 
%which do not have contextual information to adequately handle
%the exception and simply terminates the application.

%We could observe that although been known as a bad practice, the signaling general 
%exceptions in this context it was an intended design decision when no specific recovery 
%action can be defined for the exception.

%Even an instance of Throwable as signaled in one project. It points to a the
%"dont worry from know behaviour".

%\subsection{Combining Stack Trace Information with Exception Type Information}

%todo{DISCOMMENT AFTER SUBMISSION}
%\subsection{Undocumented runtime exceptions signaled by third party code}

%In this study we also classified each stack trace according to the type of the
%root exception been signaled. Table~\ref{typeroottab} presents the types of
%root causes of all stack traces reported on the Android repositories.

%TO DO: INCLUDE A NEW TABLE WITH NON JAVA.LANG RUNTIME EXCEPTIONS. Point to we cannot trust developers to document.

%This result is aligned with the results of other study conducted by from Cabral
%and Marcques ~\cite{} which also observe that runtime exceptions are most often
%not documented.

%\textbf{When libraries are used, the developer does not have access to their
%source code and thus needs to rely on the library documentation about the
%runtime exceptions that should be thrown – which, more often than not, are
%neither complete nor precise (Thomas, 2002; Sacramento et al., 2006). As a
%consequence, unhandled unchecked exceptions can be seen as one of the major
%sources of bugs in current Java systems (Jo et al., 2004). }

%On the other hand, we could observe that most of the exceptions signaled by the
%set of libraries reused by Android platform (i.e., org.apache.harmony,
%org.w3c.dom, sun.misc, org.apache.http, org.json, org.xml, javax.) referred in
%this work as libcore, were checked exceptions. It is pointed as a good practice
%for libraries (see Section ~\ref{best}) because by using checked exceptions
%libraries can define a precise exception interface to its clients. Such
%libraries are heavily used in several projects, and such precise exception
%interface may be related to the libraries maturity.

@@

In this study we also classified each stack trace according to the type of the
root exception been signaled. Table~\ref{typeroottab} presents the types of root
causes of all stack traces reported on the Android repositories.

Although most of the reported were runtime, we could also find checked
exceptions as the root causes of stack traces reported on issues. An exception
can remain uncaught in two circumstances: if all methods on the execution trace
on which this exception flows explicitly specifies the exception, or if the
checked exception is wrapped in a runtime and than re-thrown.Since checked
exceptions should be used to represent recoverable conditions, ignoring a
checked exception is considered a bad practice. When the developer is confronted
with a checked exception, the designer of the API is telling him to handle the
exceptional condition. To better understand what was causing checked exceptions
to scape from being handled we investigated the kinds of wrappings that happened
on stacks the stacks.  Next sections presents the wrappings found and discuss
about them.

%\noindent\emph{3) Odd Wrappings}
\subsection{Odd Wrappings}
Table~\ref{wrappingandroid} presents the wrappings found in this study for all
stack traces found in Android repositories. As we can see, some of the checked
exceptions where indeed wrapped in runtime exceptions or even errors, while most
of them were not wrapped along the stack trace (represented as the table lines
on which the root cause is none). This analysis also revealed unexpected
wrappings such as: Error wrapping a Checked exception; a Runtime wrapping and
Error; and an Error wrapping a Runtime.

@@

Java is the only language that provides a hybrid exception handling mechanism
which offers different types to represent different exception behaviors (see
Section ~\ref{extypes}) . According to Java specification Errors should not be
handled inside the system since they usually represent unrecoverable conditions
detected by the JVM such as OutOfMemoryError. Checked exceptions should
represent recoverable conditions and  Runtime exceptions represent are usually
related to coding errors from which the developer is not expected to recover
from.

>>>>

To guide this exploration we compiled general guidelines on how to use
exceptions proposed by Gosling~\cite{gosling2000java},
Wirfs-Brock~\cite{wirfs2006toward} and Bloch~\cite{bloch2008effective} and then
we investigate what stack traces can tell us about the adherence to such
practices in Java programs.
%~\todo{This should be our motivation, as this is
%our biggest contribution. I already changed the text. Lets discuss about this
%new version}
%RSC: I AGREE.

%GGG: I think this is too much detail for the introduction In the second phase
%of this study, we combined the information of the mined stack traces with
%source code and bytecode analysis. To do so we selected the subset of 482
%Android repositories. The Android repositories were chosen because this kind of
%application is known for dealing with several sources of exceptions, as they
%are usually io-intensive (i.e., deals with Network, Bluetooth, camera, and
%other resources), and multi-htreaded. Moreover, they are based on a single
%platform which allowed more fine-grained analysis of their exception behaviour.
%RSC: I AGREE.

@@

%GGG: Personally, I hate this paragraph in all papers as it usually reads like a space-filler
%RSC: If needed we can remove it.
The remainder of this paper is organized as follows. Section 2 presents a
background on error handling mechanisms and Java exception handling constructs
and practices. Section 3 defines research questions, the configuration of our
exploratory study, the target applications and the evaluation procedures.
Section 4 reports our analysis of the empirical data collected relating to the
first research questions. Section X is dedicated to provide insights for the 3rd
research question. Section 6 provides further discussions and lessons learned
and discusses the threats to validity associated to this study. Finally Section
7 describes the related work, and Section 8 presents our conclusions and
directions for future work.

@@

%\subsection{Error handling in programming languages}

Techniques for error detection and handling are not  an optional add-on but a
fundamental part of the system~\cite{bruntink2006discovering}. Several
approaches exists for detecting errors and responding to them, such as the return-code 
idiom (very popular in C programs and operating systems), and the exception handling 
mechanism, embedded in many main
stream programming languages.

@@

In Java, an exception is represented according to the class hierarchy shown in
Figure~\ref{fig:exchier}.  According to it every exception is an
instance of the Throwable class, and can be of three kinds: checked exceptions
(extends Exception), run-time exceptions (extends RuntimeException), and errors
(extends Error)~\cite{gosling2000java}.

@@

In Java an exception can be signaled in one of the following
circumstances~\cite{gosling2000java}: (i) explicitly thrown when a throw statement is executed; 
(ii) implicitly thrown by the JVM when the evaluation of an expression
 violates the normal semantics of language (e.g., out-of-bounds array index,
division-by-zero, access to a null reference); or (iii) implicitly thrown by
the JVM due to an internal error or resource limitation (e.g.,
OutofMemoryError).

%\todo{This sub-section is quite similar to the section introduction, shouldn't
%we merge them?}

When an exception is signaled it causes a transfer of control from the point
where the exception occurred to a point that can be specified by the programmer
(exception handler) - in Java it is represented by the try-catch block .

%and is
%responsible for performing any recovery action necessary to bring the software
%system back to a normal state and, whenever this is not possible, to log the
%exception and abort the system in an pre-defined safe way.

%In Java, as in many languages, 
The search for the handler to deal with a raised
exception occurs along the dynamic invocation chain. 
%This is claimed to increase
%the reusability, since the invoker of an operation can handle it in a wider
%context ~\cite{goodenough1975exception}. 
If no handler is found in the chain, the default handler associated to the 
Thread or ThreadGroup is invoked, which usually terminates the system 
and prints the exception stack trace.

%\subsubsection{Exception Chaining} Exception Chaining 
%This stack trace was extracted from an issue of an Android application in
%Github.

Figure~\ref{fig:wrapping} presents an example of a stack trace, on which multiple 
exception chaining is depicted. Exception Chaining (or exception wrapping) is  
common way of propagating exceptions  across components of different 
abstraction levels. It is implemented by special exception 
constructor which receives another exception as a parameter.

@@

Such undocumented runtime exceptions represent a serious threat to system robustness, specially
when such exceptions are thrown then the third party code is invoked inside the application (e.g. libraries, or framework utility code).
In such cases the client usualy do not have access to the source code, and in the absense of
the exception documentation it is very difficult or even impossible for the client of such thrid party code to 
protect the application against undocumented runtime exceptions. As a consequenve, the
 undocumented runtime exception may remain uncaught and lead to system crashes.

@@

If we filter the exceptions implicitly signaled by JVM (due to coding erros) the number of runtime
exceptions thrown by third party code is XXX signaled by Android and YYY signaled by library code.

Manually inspecting 10% of each set we could also observe that none of the methods throwing 
runtime exceptions documented the exceptions on the method signature (as suggested by Boch
 ~\cite{}) nor included a comment in Javadoc. This result is aligned with the results of other 
study conducted by from Cabral and Marcques ~\cite{} which  observed that 
runtime exceptions in .Net programs are most often not documented.

On the other hand, we could observe that the checked exceptionsmost represent most of the  
exceptions signaled by the set of libraries reused by Android platform (i.e., org.apache.harmony,
org.w3c.dom, sun.misc, org.apache.http, org.json, org.xml, javax.) referred in
this work as libcore. It is pointed as a good practice
for libraries (see Section ~\ref{best}) because by using checked exceptions
libraries can define a precise exception interface to its clients. Such
libraries are heavily used in several projects, and such precise exception
interface may be related to the libraries maturity.

@@

Although most of the  reported exceptions were runtime, we could also find checked
exceptions as the root causes of stack traces reported on issues. A checked exception
can remain uncaught in two circumstances: if all methods on the execution trace
on which this exception flows explicitly specifies the exception, or if the
checked exception is wrapped in a runtime and than re-thrown.

Since checked exceptions should be used to represent recoverable conditions, ignoring a
checked exception, is considered a bad practice. As mentioned before, when the developer is confronted
with a checked exception, the designer of the API is telling him to handle the
exceptional condition. To better understand what was causing checked exceptions
to scape from being handled we investigated the kinds of wrappings that happened
on stacks the stacks.  Next sections presents the wrappings found and discuss
about them.

%\noindent\emph{3) Odd Wrappings}
\subsection{Odd Wrappings}
Table~\ref{wrappingandroid} presents the wrappings found in this study for all
stack traces found in Android repositories. As we can see, some of the checked
exceptions where indeed wrapped in runtime exceptions or even errors, while most
of them were not wrapped along the stack trace (represented as the table lines
on which the Wapper type is -).

@@

This analysis also revealed unexpected wrappings such as: Error wrapping a Checked 
exception; a Runtime wrapping and Error; and an Error wrapping a Runtime.
Java is the only language that provides a hybrid exception handling mechanism
which offers  different types to represent different exception behaviors (i.e., error, 
runtime and checked) (see Section ~\ref{extypes}) . According to Java specification Errors should not be
handled inside the system since they usually represent unrecoverable conditions
detected by the JVM such as OutOfMemoryError. Checked exceptions should
represent recoverable conditions and  Runtime exceptions represent are usually
related to coding errors from which the developer is not expected to recover
from.
